---
title: "Appendix: R Code"
output: 
  bookdown::pdf_book:
    toc: false
    fontsize: 10pt
    includes:
      in_header: code.sty
editor_options: 
  # chunk_output_type: console
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ans<-TRUE
```

```{r}
# Import the necessary libraries
library(boot)
library(ggplot2)
library(gridExtra)
library(MASS)
if (!require("MVN")) {
  install.packages("MVN")
}
library(MVN)
```

# Load the data and extract the variables

```{r}
# Load the data
load("1854554.RData")

# Tell R that treatment is categorical
dat$treatment <- as.factor(dat$treatment)
# Add the pulse difference to the data frame
dat$pulse_diff = dat$stimulated_pulse - dat$rest_pulse

# Attach the data frame
attach(dat)
```

```{r}
# View the data
# View(dat)
```

# Exploratory analysis of the data

```{r}
# Print the first few rows of the data
head(dat)
```

```{r}
# Summary of the data
summary(dat)
```

```{r}
par(mfrow = c(1, 3)) 

# Box plot for BMI by Treatment
boxplot(bmi ~ treatment, main = "BMI",
        xlab = "Treatment", ylab = "BMI")

# Box plot for Resting Pulse by Treatment
boxplot(rest_pulse ~ treatment, main = "Resting Pulse",
        xlab = "Treatment", ylab = "Resting Pulse")

# Box plot for Stimulated Pulse by Treatment
boxplot(stimulated_pulse ~ treatment, main = "Stimulated Pulse",
        xlab = "Treatment", ylab = "Stimulated Pulse")
```

```{r}
# Plot the pulse difference
plot(pulse_diff ~ bmi, xlab = "BMI", ylab = "Pulse Difference")
```

```{r}
remove_outliers <- function(df, column_name) {
  Q1 <- quantile(df[[column_name]], 0.25)
  Q3 <- quantile(df[[column_name]], 0.75)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  df <- df[df[[column_name]] >= lower_bound & df[[column_name]] <= upper_bound, ]
  return(df)
}

# Remove the outliers
# dat <- remove_outliers(dat, "bmi")
# dat <- remove_outliers(dat, "rest_pulse")
# dat <- remove_outliers(dat, "stimulated_pulse")
# Attach the data frame
# attach(dat)
```

```{r}
# Set up the plot area
par(mfrow = c(2, 2))

# Scatter plots for the rest pulse
plot(rest_pulse[treatment == 0], stimulated_pulse[treatment == 0],
     xlab = "Rest Pulse", ylab = "Stimulated Pulse",
     main = 1, xlim = range(rest_pulse), ylim = range(stimulated_pulse))
plot(rest_pulse[treatment == 1], stimulated_pulse[treatment == 1],
     xlab = "Rest Pulse", ylab = "Stimulated Pulse",
     main = 0, xlim = range(rest_pulse), ylim = range(stimulated_pulse))

# Scatter plots for the BMI
plot(bmi[treatment == 0], stimulated_pulse[treatment == 0],
     xlab = "BMI", ylab = "Stimulated Pulse",
     main = 1, xlim = range(bmi), ylim = range(stimulated_pulse))
plot(bmi[treatment == 1], stimulated_pulse[treatment == 1],
     xlab = "BMI", ylab = "Stimulated Pulse",
     main = 0, xlim = range(bmi), ylim = range(stimulated_pulse))
```

# Initial model suggested by the clinician

```{r}
# Fit the initial linear model
fit0 <- lm(stimulated_pulse ~ rest_pulse + treatment)
```

```{r}
# Set up the plot area
par(mfrow = c(2, 2))

# Summary of the initial model
summary(fit0)

# Plot the initial model
par(mfrow = c(2, 2))
plot(fit0)
```

# Model suggested by the statistician

```{r}
custom_glm_model <-
  function(X,
           y,
           max_iter = 100,
           tol = 1e-4,
           eps = 1e-6) {
    # Initialize mu and eta
    mu <- rep(mean(y), length(y))
    eta <- 1 / mu  
    
    # Initialize beta
    beta <- rep(0, ncol(X))
    
    for (i in 1:max_iter) {
      # Update the adjusted dependent variable z
      z <- eta - ((y - mu) / mu ^ 2)
      
      # Update the weights
      w <-  mu ^ 2
      
      # Update the model using a weighted least squares fit
      WLS_fit <- lm(z ~ bmi + treatment, weights = w)
      new_beta <- as.numeric(WLS_fit$coeff)
      
      # Check if the coefficients have converged
      if (max(abs(new_beta - beta)) < tol) {
        return(new_beta)
      }
      
      # Update the coefficients
      beta <- new_beta
      
      # Update the linear predictor
      eta <- X %*% beta
      
      # Update the mean
      mu <- 1 / eta
    }
    
    return (beta)
  }

X <- cbind(1, bmi, treatment)
y <- pulse_diff

# Fit the custom GLM model
coeffs <- custom_glm_model(X, y, max_iter=1000)
coeffs
```

The algorithm converges pretty rapidly and we get similar coefficients
to the ones obtained by the R's `glm` function. Hence, from now on, we
will use the `glm` function to fit the models.

```{r}
# Fit the GLM model
fit1_form <- (pulse_diff ~ bmi + treatment)
fit1 <- glm(fit1_form, family = Gamma(link = "inverse"), data = dat)

# Summary of the GLM model
summary(fit1)

# Plot the GLM model
par(mfrow = c(2, 2))
plot(fit1)
```

# Estimation of the treatment effect

```{r}
ci_bootstrap <- function(data, fit_form, num_iterations = 1000, seed=60012) {
  # Set the seed for reproducibility
  set.seed(seed)
  
  # Compute the typical BMI (mean BMI for the sample)
  typical_bmi <- mean(bmi)
  
  # Extract the number of points in the data
  num_points <- nrow(dat)
  
  # Create an empty vector to store the predicted differences
  predicted_diff <- numeric(num_iterations)
  
  for (i in 1:num_iterations) {
    # Sample the data with replacement
    indexes <- sample(num_points, num_points, replace = TRUE)
    
    # Train the model
    fit_i <- glm(fit_form, family = Gamma(link = "inverse"), data = data[indexes,])
    
    # Predict the difference in pulse
    new_data <- data.frame(bmi = typical_bmi,
                           treatment = factor(c(0, 1),levels = levels(treatment)))
    
    predictions <- predict(fit_i, newdata = new_data, type = "response")
    predicted_diff[i] <- predictions[1] - predictions[2]
  }
  
  # Compute the estimated difference
  estimated_diff <- mean(predicted_diff)
  
  # Compute the 95% confidence interval
  conf_interval <- quantile(predicted_diff, c(0.025, 0.975))
  
  return(list(estimated_diff = estimated_diff,
              conf_interval = conf_interval))
}

ci_bootstrap(dat, fit1_form, num_iterations = 1000)
```

# Changing the linear predictors

```{r}
# Fit the GLM model
fit2 <- glm(pulse_diff ~ bmi, family = Gamma(link = "inverse"), data = dat)

# Summary of the GLM model
summary(fit2)

# Plot the GLM model
par(mfrow = c(2, 2))
plot(fit2)
```

```{r}
# Fit the GLM model
fit3 <- glm(pulse_diff ~ treatment, family = Gamma(link = "inverse"), data = dat)

# Summary of the GLM model
summary(fit3)

# Plot the GLM model
par(mfrow = c(2, 2))
plot(fit3)
```

```{r}
# Fit the GLM model
fit4_form <- (pulse_diff ~ bmi * treatment)
fit4 <- glm(fit4_form, family = Gamma(link = "inverse"), data = dat)

summary(fit4)

# Plot the GLM model
par(mfrow = c(2, 2))
plot(fit4)
```

Now let's compare the models.

```{r}
# Compare the models
anova(fit1, fit2, fit3, fit4)
AIC(fit1, fit2, fit3, fit4)
```

```{r}
# Pick the best model
fit_form <- fit4_form
fit <- fit4

# Make confidence intervals for the coefficients
confint(fit, level = 0.95)
```

# Expected increase in heart rate

```{r}
# Plot the difference in heart rate as a function of BMI
par(mfrow = c(1, 2))
plot(bmi[treatment == 1], pulse_diff[treatment == 1],
     xlab = "BMI", ylab = "Pulse Difference",
     main = 1, xlim = range(bmi), ylim = range(pulse_diff))
plot(bmi[treatment == 0], pulse_diff[treatment == 0],
     xlab = "BMI", ylab = "Pulse Difference",
     main = 0, xlim = range(bmi), ylim = range(pulse_diff))
```

```{r}
make_predictions <- function(fit, range) {
  # Prepare a data frame for prediction: repeat each BMI for both treatment levels
  data_stimulant = data.frame(bmi = range, treatment = factor(0, levels = levels(treatment)))
  data_placebo = data.frame(bmi = range, treatment = factor(1, levels = levels(treatment)))

  # Predict heart rate increase for the new data
  predictions_stimulant = predict(fit, newdata = data_stimulant, type = "response")
  predictions_placebo = predict(fit, newdata = data_placebo, type = "response")

  # Combine the prediction data frames into one, 
  # adding a column to distinguish between stimulant and placebo
  data_combined <- rbind(data_stimulant, data_placebo)
  data_combined$treatment <- factor(data_combined$treatment, labels = c(0, 1))

  # Add the predictions as a new column in the combined data frame
  data_combined$expected_increase <- c(predictions_stimulant, predictions_placebo)
  
  return (data_combined)
}

plot_predictions <- function(data_combined) {
  # Use ggplot2 to plot the expected increase in heart rate as a function of BMI
  return (ggplot(data_combined, aes(x = bmi, y = expected_increase, color = treatment)) +
    geom_line() +
    labs(title = "Expected Increase in Heart Rate",
         x = "BMI",
         y = "Expected Increase in Heart Rate") +
    scale_color_manual(values = c("blue","red")) +
    theme_minimal() +
    guides(color = guide_legend(title = "Treatment")))
}

# Make predictions
predictions_1 <- make_predictions(fit, range = seq(from = 15, to = 25, by=0.5))
predictions_2 <- make_predictions(fit, range = seq(from = 5, to = 40, by=0.5))

# Plot these values on the same graph
grid.arrange(plot_predictions(predictions_1), plot_predictions(predictions_2), ncol = 2)
```

# Simulation study to to determine the size of treatment effect

```{r}
# Set the seed
set.seed(5039)

num_simulations <- 600
treatment_effects <- seq(from = 0.1, to = 20, by = 0.1)

# Create an empty vector to store the results
power <- numeric(length(treatment_effects))

# Run the simulations
for (i in seq_along(treatment_effects)) {
  effect_size <- treatment_effects[i]
  insignificant_count <- 0
  
  # Run the simulations for the current effect size
  for (j in 1:num_simulations) {
    # Create a copy of the dataset for manipulation
    sim_data <- dat
    
    # Simulate the response variable
    sim_data$pulse_diff_sim <- unlist(simulate(fit))
    
    # Update the response variable for the placebo group 
    # (we're bringing them to the same level as the stimulant group)
    # Notice that we're adding the treatment effect to the placebo group because
    # if we would subtract it from the stimulant group, 
    # we would encounter errors with the Gamma model fitting.
    indexes <- sim_data$treatment == 1
    sim_data$pulse_diff_sim[indexes] <- sim_data$pulse_diff_sim[indexes] + effect_size
    
    # Fit the model to the simulated data
    form = pulse_diff_sim ~ bmi * treatment
    fit_sim <- glm(form, family = Gamma(link = "inverse"), data = sim_data)
    
    treatment_is_insignificant <- summary(fit_sim)$coefficients["treatment1", "Pr(>|t|)"] > 0.05
    bmi_treatment_interaction_is_insignificant <- summary(fit_sim)$coefficients["bmi:treatment1", "Pr(>|t|)"] > 0.05
    
    # Check if the treatment effect is insignificant
    if (treatment_is_insignificant && bmi_treatment_interaction_is_insignificant) {
      insignificant_count <- insignificant_count + 1
    }
  }
  
  # Store the power for the current effect size
  power[i] <- insignificant_count / num_simulations
}
```

```{r}
# Print the treatment effect after which the power is greater than 0.05
treatment_effects[which(power > 0.05)[1:10]]

# Print the treatment effect after which the power is greater than 0.6
treatment_effects[which(power > 0.6)[1:10]]

# Plot the power
par(mfrow = c(1, 1))
plot(treatment_effects, power, type = "l", xlab = "Treatment Effect", ylab = "Power", col = "blue", lwd = 2)
abline(h = 0.05, lty = 2) # Add a horizontal line at 0.05
```

# Simulation to assess the suitability of the commonly used asymptotic approximations for the sampling distribution of the maximum likelihood esimators and scaled deviance in this problem

```{r}
# Set the seed for reproducibility
set.seed(604859)

# Pick the model
model <- fit

# Set the number of simulations
num_sim <- 1000

# Extract some useful quantities
num_subjects <- nrow(dat)
num_coeffs = model$rank
dof = num_subjects - num_coeffs
coeff_names <- rownames(summary(model)$coefficients)

# Create empty matrices to store the results
coeffs_sim <- matrix(nrow = num_sim, ncol=num_coeffs)
dev_sim <- numeric(num_sim)

# Run the simulations
for (i in 1:num_sim) {
  # Simulate the output variable
  diff_sim <- unlist(simulate(model))
  
  # Fit the model to the simulated data
  form = diff_sim ~ bmi * treatment
  fit_sim <- glm(form, family = Gamma(link = "inverse"), data = dat)
  summary_fit_sim <- summary(fit_sim)
  
  # Get the dispersion parameter
  dispersion <- summary_fit_sim$dispersion
  
  # Store the coefficients 
  coeffs_sim[i,] <- coef(fit_sim)
  # Store the scaled deviance
  dev_sim[i] <- summary_fit_sim$deviance / dispersion
}
```

```{r}
# Plot the pairs of coefficients
pairs(coeffs_sim)

# Compare the residuals to the normal distribution
par(mfrow = c(2,ifelse(num_coeffs %% 2 == 0, num_coeffs / 2, (num_coeffs + 1) / 2)))
for(j in 1:num_coeffs){
  qqnorm(coeffs_sim[,j], main=paste("Normal QQ Plot -", coeff_names[j]))
  qqline(coeffs_sim[,j], col = "red")
}

# Perform the Kolmogorov-Smirnov test for the residuals
for(j in 1:num_coeffs){
  standardized_residuals <- (coeffs_sim[,j] - mean(coeffs_sim[,j])) / sd(coeffs_sim[,j])
  print(paste("Kolmogorov-Smirnov test p-value for coefficients", 
              coeff_names[j], ": ", 
              ks.test(standardized_residuals, "pnorm")$p.value))
}

# Perform a test from the MVN package in order to check the multivariate normality
mvn(coeffs_sim, mvnTest = "mardia", desc=TRUE)$multivariateNormality
```

```{r}

# Plot the histogram of simulated deviances
par(mfrow = c(1, 1))
hist(dev_sim, breaks=40, probability = TRUE, 
     main="Histogram of Simulated Scaled Deviances",
     xlab="Scaled Deviance")
# Overlay the chi-squared distribution
curve(dchisq(x, df=dof), add=TRUE, col="red", lwd=2)

# Compare the scaled deviance to the chi-squared distribution
par(mfrow = c(1, 1))
chi_sq_quantiles <- qchisq(ppoints(dev_sim), df = dof)
qqplot(chi_sq_quantiles, dev_sim)
qqline(dev_sim, col = "red")

# Perform a Kolmogorov-Smirnov test
ks.test(dev_sim, "pchisq", dof)
```

# Relationship between BMI and resting pulse

```{r}
# Print the correlation between BMI and resting pulse
cor(bmi, rest_pulse)

# Print the resting pulse as a function of BMI
par(mfrow = c(1, 2))
plot(bmi[treatment == 0], rest_pulse[treatment == 0],
     xlab = "BMI", ylab = "Resting Pulse",
     main = "Treatment", xlim = range(bmi), ylim = range(rest_pulse))
plot(bmi[treatment == 1], rest_pulse[treatment == 1],
     xlab = "BMI", ylab = "Resting Pulse",
     main = "Placebo", xlim = range(bmi), ylim = range(rest_pulse))
```
